Parameter Recovery Validation

### 2025-06-06 nbs/concepts/parameter-recovery-validation/index.qmd
docs: transition to fold-change based parameterization
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving the identifiability issues discovered in Study 2. The key innovation is the simultaneous introduction of piecewise activation dynamics and steady-state initial conditions, which together create biologically meaningful dynamics while ensuring proper parameter identifiability.

@@

The model extends Study 2 by replacing the constant transcription rate $\alpha^* = 1$ with a **piecewise constant transcription function** $\alpha^*_i(t^*)$ that varies over time for each gene. Initial conditions are fixed to the steady-state values associated with the "off" phase, eliminating the redundant parameterization between state and temporal initial conditions.

### Piecewise Transcription Rate Function

@@

$$\alpha^*_i(t^*) := \begin{cases}
\alpha^*_{\text{off}i} & \text{if } 0 \leq t^* < t^*_{0,\text{on}i}, \\
\alpha^*_{\text{on}i} & \text{if } t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i, \\
\alpha^*_{\text{off}i} & \text{if } t^*_{0,\text{on}i} + \delta^*_i \leq t^*
\end{cases}$$

where:
- $\alpha^*_{\text{off}i} = \alpha_{\text{off}i}/\beta_i$ is the dimensionless basal transcription rate
- $\alpha^*_{\text{on}i} = \alpha_{\text{on}i}/\beta_i$ is the dimensionless active transcription rate
- $t^*_{0,\text{on}i} = \beta_i \cdot t_{0,\text{on}i}$ is the dimensionless activation onset time
- $\delta^*_i = \beta_i \cdot \delta_i$ is the dimensionless activation duration

@@

To resolve identifiability, initial conditions are fixed to the steady-state values corresponding to the "off" phase:

$$u^*_{0i} = \alpha^*_{\text{off}i}, \quad s^*_{0i} = \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}$$

This eliminates the need to infer $u^*_{0i}$ and $s^*_{0i}$ as separate parameters while providing equivalent modeling flexibility through the $\alpha^*_{\text{off}i}$ parameter.

@@

#### **Phase 1: Off State** ($0 \leq t^* < t^*_{0,\text{on}i}$)

With constant transcription rate $\alpha^*_{\text{off}i}$ and initial conditions at steady state, the system remains at equilibrium:

$$\begin{align}
u^*_{ij}(t^*) &= \alpha^*_{\text{off}i} \\
s^*_{ij}(t^*) &= \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}
\end{align}$$

@@

**Initial conditions for Phase 2:**
$$u^*_{\text{on},0i} = \alpha^*_{\text{off}i}, \quad s^*_{\text{on},0i} = \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}$$

@@

For $\gamma^*_i \neq 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} \\
s^*_{ij}(\tau_{\text{on}}) &= \frac{\alpha^*_{\text{on}i}}{\gamma^*_i} + \left(\frac{\alpha^*_{\text{off}i}}{\gamma^*_i} - \xi_{\text{on}i} - \frac{\alpha^*_{\text{on}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{on}}} + \xi_{\text{on}i} e^{-\tau_{\text{on}}}
\end{align}$$

where $\xi_{\text{on}i} = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$.

For $\gamma^*_i = 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} \\
s^*_{ij}(\tau_{\text{on}}) &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\tau_{\text{on}}} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) \tau_{\text{on}} e^{-\tau_{\text{on}}}
\end{align}$$

@@

For $\gamma^*_i \neq 1$:
$$\begin{align}
u^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} \\
s^*_{\text{off},0i} &= \frac{\alpha^*_{\text{on}i}}{\gamma^*_i} + \left(\frac{\alpha^*_{\text{off}i}}{\gamma^*_i} - \xi_{\text{on}i} - \frac{\alpha^*_{\text{on}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \delta^*_i} + \xi_{\text{on}i} e^{-\delta^*_i}
\end{align}$$

For $\gamma^*_i = 1$:
$$\begin{align}
u^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} \\
s^*_{\text{off},0i} &= \alpha^*_{\text{on}i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) e^{-\delta^*_i} + (\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}) \delta^*_i e^{-\delta^*_i}
\end{align}$$

where $\xi_{\text{on}i} = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$ for the $\gamma^*_i \neq 1$ case.

@@

For $\gamma^*_i \neq 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} \\
s^*_{ij}(\tau_{\text{off}}) &= \frac{\alpha^*_{\text{off}i}}{\gamma^*_i} + \left(s^*_{\text{off},0i} - \xi_{\text{off}i} - \frac{\alpha^*_{\text{off}i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{off}}} + \xi_{\text{off}i} e^{-\tau_{\text{off}}}
\end{align}$$

where $\xi_{\text{off}i} = \frac{u^*_{\text{off},0i} - \alpha^*_{\text{off}i}}{\gamma^*_i - 1}$.

For $\gamma^*_i = 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} \\
s^*_{ij}(\tau_{\text{off}}) &= \alpha^*_{\text{off}i} + (s^*_{\text{off},0i} - \alpha^*_{\text{off}i}) e^{-\tau_{\text{off}}} + (u^*_{\text{off},0i} - \alpha^*_{\text{off}i}) \tau_{\text{off}} e^{-\tau_{\text{off}}}
\end{align}$$

@@

**Characteristics**: Genes that start with low expression and become strongly activated during the process.

@@

- $\alpha^*_{\text{off}i} < 0.2$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.5$ (strong activation)
- $t^*_{0,\text{on}i} < 0.4$ (early activation onset)
- $\delta^*_i > 0.3$ (sustained activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 7.5$ (strong upregulation)

@@

#### **2. Decay-Only Patterns**

@@

- $\alpha^*_{\text{off}i} > 0.5$ (high basal transcription)
- $t^*_{0,\text{on}i} > T^*_M$ (activation onset beyond observation window)
- $\delta^*_i$ (irrelevant, no activation observed)
- **Effective dynamics**: Monotonic decay from high initial levels

**Biological examples**: Maternal transcripts during development, housekeeping genes during stress, pluripotency factors during differentiation

#### **3. Transient Activation Patterns**
**Characteristics**: Genes that show brief, pulse-like activation followed by return to baseline.

@@

- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.0$ (moderate to strong activation)
- $t^*_{0,\text{on}i} < 0.5$ (early to mid-process activation)
- $\delta^*_i < 0.4$ (brief activation duration) *[optimized from < 0.3]*
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (clear but not extreme upregulation)

@@

#### **4. Sustained Activation Patterns**
**Characteristics**: Genes that become activated and remain highly expressed throughout most of the process.

@@

- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.0$ (strong activation)
- $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
- $\delta^*_i > 0.45$ (long activation duration) *[optimized from > 0.6]*
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (sustained upregulation)

@@

@@

The pattern constraints above were optimized using a systematic CDF-based analysis to ensure mathematical feasibility while preserving biological interpretability. This section documents the optimization process and provides transparency about how we arrived at the final parameter specifications.

@@

For each pattern constraint, we calculate the probability that a parameter sampled from its prior distribution satisfies the constraint using cumulative distribution functions (CDFs). For a LogNormal prior $X \sim \text{LogNormal}(\mu, \sigma^2)$, the probability of satisfying constraint $X < \theta$ is:

$$P(X < \theta) = \Phi\left(\frac{\log(\theta) - \mu}{\sigma}\right)$$

@@

# Prior hyperparameters (optimized)
priors = {
    'alpha_off': {'loc': -2.3, 'scale': 0.5},    # log(0.1)
    'alpha_on': {'loc': 0.69, 'scale': 0.5},     # log(2.0)
    't_on_star': {'loc': -1.2, 'scale': 0.3},    # log(0.3)
    'delta_star': {'loc': -1.0, 'scale': 0.35},  # log(0.37) [optimized]
}

# Calculate constraint feasibility for each pattern
patterns = {
    'activation': [
        ('alpha_off', '<', 0.2), ('alpha_on', '>', 1.5),
        ('t_on_star', '<', 0.4), ('delta_star', '>', 0.3)
    ],
    'transient': [
        ('alpha_off', '<', 0.3), ('alpha_on', '>', 1.0),
        ('t_on_star', '<', 0.5), ('delta_star', '<', 0.4)  # optimized
    ],
    'sustained': [
        ('alpha_off', '<', 0.3), ('alpha_on', '>', 1.0),
        ('t_on_star', '<', 0.3), ('delta_star', '>', 0.45)  # optimized
    ]
}

@@

    for param, op, threshold in constraints:
        prior = priors[param]
        prob = calculate_lognormal_cdf_probability(
            prior['loc'], prior['scale'], threshold, upper=(op == '>')
        )
        joint_prob *= prob
        print(f"  P({param} {op} {threshold}) = {prob:.3f}")

@@

The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while maintaining the hierarchical latent time structure. For visual clarity, we use $\theta_i$ to represent the vector of piecewise activation parameters $(\gamma^*_i, \alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i})$ and $\delta_i$ to represent the activation duration parameter.

@@

In the PGM above, we use simplified notation for visual clarity:
- $\theta_i$ represents the vector of piecewise activation parameters $(\gamma^*_i, \alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i})$
- $\delta_i$ represents the activation duration parameter $\delta^*_i$ (waiting time from onset to deactivation)
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\delta}, \sigma_{\delta}^2$ represent the hyperparameters for $\delta_i$

**Detailed priors for piecewise activation parameters:**

The priors are designed to encode biologically reasonable parameter ranges while maintaining sufficient flexibility for diverse gene expression patterns:

\begin{align}
    \gamma^*_i &\sim \text{LogNormal}(\log(1.0), 0.5^2) \quad \text{(relative degradation rate)} \\
    \alpha^*_{\text{off}i} &\sim \text{LogNormal}(\log(0.1), 0.5^2) \quad \text{(basal transcription rate)} \\
    \alpha^*_{\text{on}i} &\sim \text{LogNormal}(\log(2.0), 0.5^2) \quad \text{(active transcription rate)} \\
    t^*_{0,\text{on}i} &\sim \text{LogNormal}(\log(0.3), 0.3^2) \quad \text{(activation onset time)} \\
    \delta^*_i &\sim \text{LogNormal}(\log(0.37), 0.35^2) \quad \text{(activation duration, optimized)}
\end{align}

@@

- **$\alpha^*_{\text{off}i} \sim \text{LogNormal}(\log(0.1), 0.5^2)$**: Centers basal transcription at 10% of characteristic rate, with 95% between ~0.037 and ~0.27. This represents low but non-zero basal expression typical of repressed genes.

- **$\alpha^*_{\text{on}i} \sim \text{LogNormal}(\log(2.0), 0.5^2)$**: Centers active transcription at 2× characteristic rate, with 95% between ~0.74 and ~5.4. This ensures activation represents meaningful upregulation while avoiding extreme values.

- **$t^*_{0,\text{on}i} \sim \text{LogNormal}(\log(0.3), 0.3^2)$**: Centers activation onset at 30% through the process timeline, with 95% between ~0.17 and ~0.53. This favors early activation while allowing variation in timing.

- **$\delta^*_i \sim \text{LogNormal}(\log(0.37), 0.35^2)$**: Centers activation duration at 37% of the process timeline, with 95% between ~0.18 and ~0.76. This distribution was optimized to support both transient (< 40%) and sustained (> 45%) activation patterns with sufficient feasibility.

@@

#### **Identifiability Resolution**

- **Eliminates redundant parameterization**: No separate inference of $u^*_{0i}, s^*_{0i}$ parameters
- **Biological realism**: Steady-state initial conditions are more plausible than arbitrary values
- **Clear parameter roles**: Each piecewise parameter has distinct biological interpretation

@@

- **Activation genes**: Transition from low to high expression
- **Decay genes**: Start high, remain high (no activation within observation window)
- **Transient genes**: Brief activation followed by return to baseline
- **Sustained genes**: Long-duration activation

@@

This study will validate the model's ability to:

1. **Recover piecewise activation parameters**: $\alpha^*_{\text{off}i}, \alpha^*_{\text{on}i}, t^*_{0,\text{on}i}, \delta^*_i$
2. **Distinguish gene expression patterns**: Classify genes as activation, decay, transient, or sustained
3. **Infer hierarchical time structure**: Recover population-level temporal parameters
4. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
5. **Scale to multiple genes**: Demonstrate efficiency with gene panels of varying sizes
6. **Numerical robustness**: Maintain stability across the $\gamma^*_i = 1$ boundary
7. **Phase assignment accuracy**: Correctly identify which phase each cell-gene pair occupies

@@

# Validate prior parameter ranges
print("\nPrior parameter range validation:")
for param_name, samples in prior_parameter_samples.items():
    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
        print(f"{param_name}:")
        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
```

>>>>

This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving critical dimensional analysis inconsistencies identified in the piecewise activation framework. The key innovation is a **dimensionally consistent parameterization** that eliminates parameter redundancy by fixing the reference transcription state, while maintaining biological interpretability through fold-change parameters.

@@

The model extends Study 2 by replacing the constant transcription rate $\alpha^* = 1$ with a **piecewise constant transcription function** $\alpha^*_i(t^*)$ that varies over time for each gene. To resolve dimensional analysis violations, we implement a **constrained parameterization** where the off-state transcription rate serves as a fixed reference ($\alpha^*_{\text{off}i} = 1.0$), eliminating parameter redundancy while maintaining biological interpretability through fold-change parameters.

### Dimensionally Consistent Piecewise Framework

#### Reference State Parameterization

To eliminate parameter redundancy identified in the dimensional analysis, we fix the off-state as a reference:

$$\alpha^*_{\text{off}i} = 1.0 \quad \text{(fixed reference, not inferred)}$$

This choice establishes a **consistent dimensional reference** where $U_{0i}$ unambiguously represents the characteristic concentration scale corresponding to the off-state transcription rate.

#### Extended Temporal Framework with Pre-Activation

To accommodate realistic gene expression scenarios, including genes that are already activated when observation begins, we extend the temporal framework to allow **negative onset times**.

#### Piecewise Transcription Rate Function

@@

$$\alpha^*_i(t^*) := \begin{cases}
1.0 & \text{if } t^* < t^*_{0,\text{on}i}, \\
R_{\text{on},i} & \text{if } t^*_{0,\text{on}i} \leq t^* < t^*_{0,\text{on}i} + \delta^*_i, \\
1.0 & \text{if } t^* \geq t^*_{0,\text{on}i} + \delta^*_i
\end{cases}$$

where:
- $\alpha^*_{\text{off}i} = 1.0$ is the fixed reference transcription rate
- $R_{\text{on},i} = \alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} = \alpha^*_{\text{on}i}$ is the fold-change during activation (inferred parameter)
- $t^*_{0,\text{on}i}$ is the dimensionless activation onset time (can be negative for pre-activation scenarios)
- $\delta^*_i$ is the dimensionless activation duration

#### Biological Interpretation of Negative Onset Times

When $t^*_{0,\text{on}i} < 0$:

- The gene was **already activated** before observation began ($t^* = 0$)
- Cells at early timepoints may show **high expression** for such genes
- The model captures **pre-existing activation states** in addition to activation events during observation
- **Biological examples**: Early developmental genes already active when profiling begins, stress response genes in cells experiencing ongoing stress, lineage commitment factors in partially committed progenitors

@@

Initial conditions are fixed to the off-state steady-state values:

$$u^*_{0i} = 1.0, \quad s^*_{0i} = \frac{1.0}{\gamma^*_i} = \frac{1}{\gamma^*_i}$$

This eliminates the need to infer $u^*_{0i}$ and $s^*_{0i}$ as separate parameters while providing equivalent modeling flexibility through the characteristic scale $U_{0i}$ and fold-change $R_{\text{on},i}$.

@@

#### **Phase 1: Off State** ($t^* < t^*_{0,\text{on}i}$)

With constant transcription rate $\alpha^*_{\text{off}i} = 1.0$ and initial conditions at steady state, the system remains at equilibrium:

$$\begin{align}
u^*_{ij}(t^*) &= 1.0 \\
s^*_{ij}(t^*) &= \frac{1.0}{\gamma^*_i}
\end{align}$$

**Note**: This phase may not be observed if $t^*_{0,\text{on}i} < 0$ (pre-activation scenario).

@@

**Initial conditions for Phase 2:**
$$u^*_{\text{on},0i} = 1.0, \quad s^*_{\text{on},0i} = \frac{1.0}{\gamma^*_i}$$

@@

For $\gamma^*_i \neq 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{on}}) &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\tau_{\text{on}}} \\
s^*_{ij}(\tau_{\text{on}}) &= \frac{R_{\text{on},i}}{\gamma^*_i} + \left(\frac{1.0}{\gamma^*_i} - \xi_{\text{on}i} - \frac{R_{\text{on},i}}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{on}}} + \xi_{\text{on}i} e^{-\tau_{\text{on}}}
\end{align}$$

where $\xi_{\text{on}i} = \frac{1.0 - R_{\text{on},i}}{\gamma^*_i - 1}$.

For $\gamma^*_i = 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{on}}) &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\tau_{\text{on}}} \\
s^*_{ij}(\tau_{\text{on}}) &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\tau_{\text{on}}} + (1.0 - R_{\text{on},i}) \tau_{\text{on}} e^{-\tau_{\text{on}}}
\end{align}$$

@@

For $\gamma^*_i \neq 1$:
$$\begin{align}
u^*_{\text{off},0i} &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\delta^*_i} \\
s^*_{\text{off},0i} &= \frac{R_{\text{on},i}}{\gamma^*_i} + \left(\frac{1.0}{\gamma^*_i} - \xi_{\text{on}i} - \frac{R_{\text{on},i}}{\gamma^*_i}\right) e^{-\gamma^*_i \delta^*_i} + \xi_{\text{on}i} e^{-\delta^*_i}
\end{align}$$

For $\gamma^*_i = 1$:
$$\begin{align}
u^*_{\text{off},0i} &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\delta^*_i} \\
s^*_{\text{off},0i} &= R_{\text{on},i} + (1.0 - R_{\text{on},i}) e^{-\delta^*_i} + (1.0 - R_{\text{on},i}) \delta^*_i e^{-\delta^*_i}
\end{align}$$

where $\xi_{\text{on}i} = \frac{1.0 - R_{\text{on},i}}{\gamma^*_i - 1}$ for the $\gamma^*_i \neq 1$ case.

@@

For $\gamma^*_i \neq 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{off}}) &= 1.0 + (u^*_{\text{off},0i} - 1.0) e^{-\tau_{\text{off}}} \\
s^*_{ij}(\tau_{\text{off}}) &= \frac{1.0}{\gamma^*_i} + \left(s^*_{\text{off},0i} - \xi_{\text{off}i} - \frac{1.0}{\gamma^*_i}\right) e^{-\gamma^*_i \tau_{\text{off}}} + \xi_{\text{off}i} e^{-\tau_{\text{off}}}
\end{align}$$

where $\xi_{\text{off}i} = \frac{u^*_{\text{off},0i} - 1.0}{\gamma^*_i - 1}$.

For $\gamma^*_i = 1$:
$$\begin{align}
u^*_{ij}(\tau_{\text{off}}) &= 1.0 + (u^*_{\text{off},0i} - 1.0) e^{-\tau_{\text{off}}} \\
s^*_{ij}(\tau_{\text{off}}) &= 1.0 + (s^*_{\text{off},0i} - 1.0) e^{-\tau_{\text{off}}} + (u^*_{\text{off},0i} - 1.0) \tau_{\text{off}} e^{-\tau_{\text{off}}}
\end{align}$$

@@

**Characteristics**: Genes that start with baseline expression and become strongly activated during the process.

@@

- $R_{\text{on},i} > 3.0$ (strong fold-change during activation)
- $t^*_{0,\text{on}i} > 0$ and $t^*_{0,\text{on}i} < 0.4$ (activation occurs after observation start but early)
- $\delta^*_i > 0.3$ (sustained activation duration)

@@

#### **2. Pre-Activation Patterns**

**Characteristics**: Genes that were already activated before observation began, showing high initial expression.

- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change)
- $t^*_{0,\text{on}i} < 0$ (activation onset before observation window)
- $\delta^*_i > |t^*_{0,\text{on}i}|$ (activation duration extends into observation window)

**Biological examples**: Early developmental genes, lineage commitment factors, stress response genes in ongoing stress

#### **3. Decay-Only Patterns**

@@

- $R_{\text{on},i} > 1.5$ (moderate fold-change, but not observed during activation)
- $t^*_{0,\text{on}i} > T^*_M$ (activation onset beyond observation window) OR
- $t^*_{0,\text{on}i} < 0$ and $\delta^*_i < |t^*_{0,\text{on}i}|$ (activation ended before observation)

**Biological examples**: Maternal transcripts during development, pluripotency factors during differentiation

#### **4. Transient Activation Patterns**
**Characteristics**: Genes that show brief, pulse-like activation followed by return to baseline.

@@

- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change)
- $t^*_{0,\text{on}i} > 0$ and $t^*_{0,\text{on}i} < 0.5$ (activation during observation)
- $\delta^*_i < 0.4$ (brief activation duration)

@@

#### **5. Sustained Activation Patterns**
**Characteristics**: Genes that become activated and remain highly expressed throughout most of the process.

@@

- $R_{\text{on},i} > 2.0$ (strong fold-change)
- $t^*_{0,\text{on}i} > 0$ and $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
- $\delta^*_i > 0.5$ (long activation duration)

@@

### Mathematical Framework for Pre-Activation Scenarios

The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when observation begins. This extension is mathematically valid and biologically realistic:

#### **Mathematical Validity of Negative Onset Times**

When $t^*_{0,\text{on}i} < 0$:

1. **Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$

2. **Analytical solutions remain valid**: All exponential solutions are well-defined for negative onset times

3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the observation window

#### **Implementation Considerations**

- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no observed cells will be in Phase 1
- **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the observation window

@@

The pattern constraints above were optimized using a systematic CDF-based analysis to ensure mathematical feasibility while preserving biological interpretability. The analysis now includes both LogNormal and Normal distributions to handle the extended temporal framework.

@@

For each pattern constraint, we calculate the probability that a parameter sampled from its prior distribution satisfies the constraint using cumulative distribution functions (CDFs).

For **LogNormal parameters** ($R_{\text{on},i}$, $\delta^*_i$), the probability of satisfying constraint $X > \theta$ is:

$$P(X > \theta) = 1 - \Phi\left(\frac{\log(\theta) - \mu}{\sigma}\right)$$

For **Normal parameters** ($t^*_{0,\text{on}i}$), the probability of satisfying constraint $X > \theta$ is:

$$P(X > \theta) = 1 - \Phi\left(\frac{\theta - \mu}{\sigma}\right)$$

@@

def calculate_normal_cdf_probability(loc, scale, threshold, upper=True):
    """Calculate P(X > threshold) or P(X < threshold) for Normal(loc, scale)."""
    if upper:
        return 1 - stats.norm.cdf(threshold, loc=loc, scale=scale)
    else:
        return stats.norm.cdf(threshold, loc=loc, scale=scale)

# Prior hyperparameters (updated for constrained parameterization with Normal onset times)
priors = {
    'R_on': {'loc': 0.916, 'scale': 0.4},        # log(2.5), fold-change (LogNormal)
    't_on_star': {'loc': 0.2, 'scale': 0.6},     # Normal(0.2, 0.6²), allows negatives
    'delta_star': {'loc': -1.0, 'scale': 0.35},  # log(0.37) [optimized] (LogNormal)
}

# Calculate constraint feasibility for each pattern
patterns = {
    'activation': [
        ('R_on', '>', 3.0), ('t_on_star', '>', 0.0),
        ('t_on_star', '<', 0.4), ('delta_star', '>', 0.3)
    ],
    'pre_activation': [
        ('R_on', '>', 2.0), ('t_on_star', '<', 0.0),
        ('delta_star', '>', 0.3)  # Simplified constraint
    ],
    'transient': [
        ('R_on', '>', 2.0), ('t_on_star', '>', 0.0),
        ('t_on_star', '<', 0.5), ('delta_star', '<', 0.4)
    ],
    'sustained': [
        ('R_on', '>', 2.0), ('t_on_star', '>', 0.0),
        ('t_on_star', '<', 0.3), ('delta_star', '>', 0.5)
    ]
}

@@

    for param, op, threshold in constraints:
        prior = priors[param]

        if param == 't_on_star':  # Normal distribution
            prob = calculate_normal_cdf_probability(
                prior['loc'], prior['scale'], threshold, upper=(op == '>')
            )
        else:  # LogNormal distribution
            prob = calculate_lognormal_cdf_probability(
                prior['loc'], prior['scale'], threshold, upper=(op == '>')
            )

        joint_prob *= prob
        print(f"  P({param} {op} {threshold}) = {prob:.3f}")

@@

The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while maintaining the hierarchical latent time structure. For visual clarity, we use $\theta_i$ to represent the vector of piecewise activation parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i})$ and $\delta_i$ to represent the activation duration parameter. Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.

@@

In the PGM above, we use simplified notation for visual clarity:
- $\theta_i$ represents the vector of piecewise activation parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i})$
- $\delta_i$ represents the activation duration parameter $\delta^*_i$ (waiting time from onset to deactivation)
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\delta}, \sigma_{\delta}^2$ represent the hyperparameters for $\delta_i$
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable

**Detailed priors for constrained piecewise activation parameters:**

The priors are designed to encode biologically reasonable parameter ranges while eliminating dimensional redundancy:

\begin{align}
    \gamma^*_i &\sim \text{LogNormal}(\log(1.0), 0.5^2) \quad \text{(relative degradation rate)} \\
    R_{\text{on},i} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \quad \text{(activation fold-change)} \\
    t^*_{0,\text{on}i} &\sim \text{Normal}(0.2, 0.6^2) \quad \text{(activation onset time, allows negatives)} \\
    \delta^*_i &\sim \text{LogNormal}(\log(0.37), 0.35^2) \quad \text{(activation duration, optimized)}
\end{align}

@@

- **$R_{\text{on},i} \sim \text{LogNormal}(\log(2.5), 0.4^2)$**: Centers fold-change at 2.5× during activation, with 95% between ~1.5 and ~4.2. This represents meaningful upregulation while avoiding extreme values that could cause numerical issues.

- **$t^*_{0,\text{on}i} \sim \text{Normal}(0.2, 0.6^2)$**: Centers activation onset at 20% through the process timeline, with 95% between approximately -1.0 and +1.4. **Negative values** represent genes already activated before observation began, enabling modeling of pre-activation scenarios.

- **$\delta^*_i \sim \text{LogNormal}(\log(0.37), 0.35^2)$**: Centers activation duration at 37% of the process timeline, with 95% between ~0.18 and ~0.76. This distribution was optimized to support both transient (< 40%) and sustained (> 50%) activation patterns with sufficient feasibility.

@@

#### **Resolved Dimensional Analysis Issues**

- **Eliminates parameter redundancy**: No ambiguity between $U_{0i}$ and $\alpha^*_{\text{off}i}$ since $\alpha^*_{\text{off}i} = 1.0$ (fixed)
- **Clear dimensional structure**: $U_{0i}$ unambiguously represents off-state concentration scale
- **Proper parameter hierarchy**: Each inferred parameter has a unique biological role
- **No separate inference of initial conditions**: $u^*_{0i}, s^*_{0i}$ fixed to steady-state values
- **Biological realism**: Steady-state initial conditions are more plausible than arbitrary values

#### **Enhanced Temporal Modeling**

- **Pre-activation scenarios**: Models genes already activated when observation begins
- **Flexible onset timing**: Normal prior allows negative onset times naturally
- **Biological realism**: Captures realistic experimental scenarios better
- **Maintains population structure**: Hierarchical time modeling preserved

@@

- **Activation genes**: Transition from baseline to high expression during observation
- **Pre-activation genes**: Already activated before observation, showing high initial expression
- **Decay genes**: Start high, remain high (no activation within observation window)
- **Transient genes**: Brief activation followed by return to baseline
- **Sustained genes**: Long-duration activation

@@

This study will validate the constrained model's ability to:

1. **Recover constrained piecewise parameters**: $\gamma^*_i$, $R_{\text{on},i}$, $t^*_{0,\text{on}i}$, $\delta^*_i$
2. **Handle pre-activation scenarios**: Correctly infer negative onset times and their biological implications
3. **Distinguish enhanced expression patterns**: Classify genes across all five patterns including pre-activation
4. **Maintain hierarchical time inference**: Recover population-level temporal parameters
5. **Handle realistic noise levels**: Maintain accuracy with Poisson observation noise
6. **Scale efficiently**: Demonstrate computational advantages of reduced parameter space
7. **Validate fold-change interpretation**: Ensure $R_{\text{on},i}$ estimates match true activation strength
8. **Assess dimensional consistency**: Verify elimination of parameter redundancy issues
9. **Numerical robustness**: Maintain stability across the $\gamma^*_i = 1$ boundary

@@

# Validate prior parameter ranges
print("\nPrior parameter range validation:")
for param_name, samples in prior_parameter_samples.items():
    if param_name.startswith(('R_on', 't_on', 'delta', 'gamma_star')):
        print(f"{param_name}:")
        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")
```
